Sample variance in weak lensing: how many simulations are required?

### 2016-01-23 Paper/paper.tex
Update paper text with new comments for Figure 5

Constraining cosmology using weak gravitational lensing consists of
comparing a measured feature vector of dimension $N_b$ with its
simulated counterpart. An accurate estimate of the $N_b\times N_b$
feature covariance matrix $\bb{C}$ is essential to obtain accurate
parameter confidence intervals. When $\bb{C}$ is measured from a set
of simulations, an important question is how large this set should
be. To answer this question, we construct different ensembles of $N_r$
realizations of the shear field, using a common randomization
procedure that recycles the outputs from a smaller number $N_s\leq
N_r$ of independent ray-tracing $N$--body simulations.  We study
parameter confidence intervals as a function of ($N_s,N_r$) in the
range $1\leq N_s\leq 200$ and $1\leq N_r\lesssim 10^5$.  Previous work
\citep{DodelsonSchneider13} has shown that Gaussian noise in the
feature vectors (from which the covariance is estimated) lead, at quadratic order, to an $O(1/N_r)$
degradation of the parameter confidence intervals. Using a variety of
lensing features measured in our simulations, including shear-shear
power spectra and peak counts, we show that cubic and quartic
covariance fluctuations lead to additional $O(1/N_r^2)$ error
degradation that is not negligible when $N_r$ is only a factor of few
larger than $N_b$. We study the large $N_r$ limit, and find that a
single, 240Mpc$/h$ sized, $N$--body simulation ($N_s=1$) can be repeatedly recycled to
produce as many as $N_r={\rm few}\times10^4$ shear maps whose power spectra can be treated as statistically independent. As a result, a small number of simulations ($N_s=1$ or $2$) is sufficient to forecast parameter confidence intervals at percent accuracy.

@@

\keywords{Weak Gravitational Lensing --- Simulations --- Methods: numerical,statistical}
\pacs{98.80.-k, 95.36.+x, 95.30.Sf, 98.62.Sb}

@@

We assume a uniform galaxy distribution at a constant redshift $z_s=2$
(at which the simulation box has an angular size of $\theta_{\rm
  box}=3.5\mathrm{deg}$) and we discretize the mass distribution
between $z_s$ and the observer at $z=0$ with a sequence of 46 two
dimensional lenses of thickness $80\mathrm{Mpc}/h$. The surface
density on each lens plane is computed by projecting the
three--dimensional density measured from \texttt{Gadget2}
snapshots. We then apply the multi--lens--plane algorithm (see
\citep{RayTracingHartlap,RayTracingJain} for example) to trace the
deflections of $n_{\rm ray}^2=2048^2$ light rays arranged on a square grid
of total size $\theta_{\rm box}$, from $z=0$ to $z_s$. This corresponds to a pixel angular resolution of $0.1^\prime$. 
Our implementation of this algorithm is part of the \texttt{LensTools}
computing package we have been developing \citep{LensTools}, and have
released under the \texttt{MIT} license. Many different realizations
$r$ of the same shear field $\pmb{\gamma}_r(\pmb{\theta})$ can be
generated by picking different lens planes that lie between the
observer and $z_s$. The randomization procedure we adopt is the
following (see \citep{Sato12} for reference):

@@

\begin{figure}
\includegraphics[scale=0.3]{Figures/scaling_nr.eps}
\caption{Bias in the variance of $w$,
  $\langle\h{\sigma}^2_w\rangle-\sigma^2_{w,\infty}$, as a function of
  the number of realizations $N_r$ used to estimate the covariance
  (eq.~\ref{covest}). The figure shows both the trend measured in the
  simulations (solid lines) and their scaling expected from
  Eq.~(\ref{ourscaling}) with $D=N_b-N_p$ (dashed line). The
  asymptotic variance $\sigma^2_{w,\infty}$ has been estimated to be the value $\langle\h{\sigma}^2_w\rangle(N_r=10^5)$. Different features are considered: power spectra with logarithmically spaced $\ell\in[100,6000]$ (black), $\ell\in[100,250]$ (red) and peak counts of height $\kappa_0\in[0.44,0.48]$ (blue). No deviations from the expected $1/N_r$ behavior are observed up to $N_r\approx{\rm few}\times10^4$, except for the large scale power spectrum, in which case the deviations occur much earlier ($N_r\approx 10^3$).}
\label{wvar_nr}
\end{figure}

@@

In Figure \ref{wvar_nr}, we examine how the degradation in the $w$
constraint depends on the number of simulations used to estimate the
covariance, in the limit of large $N_r$. We find excellent agreement
with the expected scaling (eq.~\ref{ourscaling}) up to $N_r\sim {\rm few}\times10^4$ when using the $\kappa$ power spectrum in the multipole range $\ell\in[100,6000]$. As the figure shows, around these values of $N_r$ the $\langle\hat{\sigma}^2_w\rangle - \sigma^2_{w,\infty}$ curve
becomes noisy and reaches negative values. This is a clear indication that the $1/N_r$ behavior is broken and a plateau in
$\langle\hat{\sigma}^2_w\rangle$ is reached. The negative values in
the plot are a consequence of the noise in the estimation of this
plateau value (or equivalently in the estimated value of
$\sigma^2_{w,\infty}$). We therefore conclude that a single $N$--body
simulation is sufficient to construct an ensemble of up to a
few$\times 10^4$ mutually independent convergence power spectra. For
$N_r\gg10^4$, the shear realizations can no longer be considered
independent. We should emphasize that the validity of this statement is conditional to the size of the simulation box (which in our case has been chosen to be 240Mpc$/h$) and the range of multipoles $\ell$ used to constrain the parameters. Figure \ref{wvar_nr} shows indeed that, when we try infer $w$ only from large scale modes, $\ell\lesssim250$, the plateau is reached at least an order of magnitude earlier in the number of realizations. In other words, the number of independent power spectra gets lower and lower as we increase the spatial scales of interest. This is due to the fact that, because of the finite box size, the number of independent lens plane shifts (as described in \S~\ref{shearsim}) decreases as the mode size approaches the size of the box.

@@

\item When the feature covariance matrix is measured from simulations,
  parameter constraints are degraded. This degradation is appreciably
  larger than the $O(1/N_r)$ computed by \citep{DodelsonSchneider13}
  when the number of realizations $N_r$ is only a factor of few larger
  than the feature vector size $N_b$.
\item We can recycle a single 240Mpc$/h$ $N$--body simulation to produce an
  ensemble of $O(10^4)$ shear maps whose power spectra are statistically independent in the range $\ell\in[100,6000]$. The mean feature measured
  from a shear ensemble, though, could be inaccurate if only one $N$--body simulation is used.
\item As few as one or two independent $N$--body simulations are
  sufficient to forecast $w$ error bars to 1\% accuracy, provided that
  a sufficiently large number $N_r$ of realizations are used to
  measure feature covariances.  In particular, provided that biases
  in the inverse covariance are corrected, percent--level forecasts
  require $N_r\gsim 100 (N_b-N_p)$ realizations.
\item Depending on the feature type used to constrain cosmology, a
  larger number of $N$--body simulations might be needed to measure
  accurate ensemble means to an accuracy corresponding 10\% of the
  statistical error. If this accuracy requirement is relaxed to 50\%
  of the statistical error, we find that as low as $N_s=2$ simulations
  are sufficient for the feature types we consider in this work.
\end{itemize}
%
Future extensions of this work should involve extending our analysis
to a larger set of cosmological parameters, and to more general
feature spaces, such as the ones that characterize non--Gaussian
statistics (e.g. including higher moments of the $\kappa$ field,
Minkowski Functionals, and higher-order $\kappa$ correlators). While
our results are highly encouraging, and suggest that a single $N$--body
simulation can be recycled repeatedly, to produce as many as $10^4$
independent shear power spectra or peak count histograms. In order to
scale our results to large future surveys, such as LSST, it will be
necessary to determine if our findings hold when challenged by larger
and higher-resolution $N$--body simulations \citep{Qcontinuum}.

>>>>

Constraining cosmology using weak gravitational lensing consists of
comparing a measured feature vector of dimension $N_b$ with its
simulated counterpart. An accurate estimate of the $N_b\times N_b$
feature covariance matrix $\bb{C}$ is essential to obtain accurate
parameter confidence intervals. When $\bb{C}$ is measured from a set
of simulations, an important question is how large this set should
be. To answer this question, we construct different ensembles of $N_r$
realizations of the shear field, using a common randomization
procedure that recycles the outputs from a smaller number $N_s\leq
N_r$ of independent ray-tracing $N$--body simulations.  We study
parameter confidence intervals as a function of ($N_s,N_r$) in the
range $1\leq N_s\leq 200$ and $1\leq N_r\lesssim 10^5$.  Previous work
\citep{DodelsonSchneider13} has shown that Gaussian noise in the
feature vectors (from which the covariance is estimated) lead, at quadratic order, to an $O(1/N_r)$
degradation of the parameter confidence intervals. Using a variety of
lensing features measured in our simulations, including shear-shear
power spectra and peak counts, we show that cubic and quartic
covariance fluctuations lead to additional $O(1/N_r^2)$ error
degradation that is not negligible when $N_r$ is only a factor of few
larger than $N_b$. We study the large $N_r$ limit, and find that a
single, 240Mpc$/h$ sized, $N$--body simulation ($N_s=1$) can be repeatedly recycled to
produce as many as $N_r={\rm few}\times10^4$ shear maps whose power spectra and high significance peak counts can be treated as statistically independent. As a result, a small number of simulations ($N_s=1$ or $2$) is sufficient to forecast parameter confidence intervals at percent accuracy.

@@

\keywords{Weak Gravitational Lensing --- Simulations --- Methods: analytical, numerical, statistical}
\pacs{98.80.-k, 95.36.+x, 95.30.Sf, 98.62.Sb}

@@

We assume a uniform galaxy distribution at a constant redshift $z_s=2$
(at which the simulation box has an angular size of $\theta_{\rm
  box}=3.5^\circ$) and we discretize the mass distribution
between $z_s$ and the observer at $z=0$ with a sequence of 46 two
dimensional lenses of thickness $80\mathrm{Mpc}/h$. The surface
density on each lens plane is computed by projecting the
three--dimensional density measured from \texttt{Gadget2}
snapshots. We then apply the multi--lens--plane algorithm (see
\citep{RayTracingHartlap,RayTracingJain} for example) to trace the
deflections of $n_{\rm ray}^2=2048^2$ light rays arranged on a square grid
of total size $\theta_{\rm box}$, from $z=0$ to $z_s$. This corresponds to a pixel angular resolution of $0.1^\prime$. 
Our implementation of this algorithm is part of the \texttt{LensTools}
computing package we have been developing \citep{LensTools}, and have
released under the \texttt{MIT} license. Many different realizations
$r$ of the same shear field $\pmb{\gamma}_r(\pmb{\theta})$ can be
generated by picking different lens planes that lie between the
observer and $z_s$. The randomization procedure we adopt is the
following (see \citep{Sato12} for reference):

@@

\begin{figure}
\includegraphics[scale=0.3]{Figures/scaling_nr.eps}
\caption{Bias in the variance of $w$,
  $\langle\h{\sigma}^2_w\rangle-\sigma^2_{w,\infty}$, as a function of
  the number of realizations $N_r$ used to estimate the covariance
  (eq.~\ref{covest}). The figure shows both the trend measured in the
  simulations (solid lines) and their scaling expected from
  Eq.~(\ref{ourscaling}) with $D=N_b-N_p$ (dashed line). The
  asymptotic variance $\sigma^2_{w,\infty}$ has been estimated to be the value $\langle\h{\sigma}^2_w\rangle(N_r=10^5)$. Different features are considered: power spectra with logarithmically spaced $\ell\in[100,6000]$ (black), $\ell\in[100,250]$ (red), peak counts in the unsmoothed maps with height $\kappa_0\in[0.44,0.48]$ (green) and peak counts in the smoothed maps (with a Gaussian kernel of size $\theta_G=1^\prime$) with height $\kappa_0>0.15$ (blue). No deviations from the expected $1/N_r$ behavior are observed up to $N_r\approx{\rm few}\times10^4$, except for the large scale power spectrum, in which case the deviations occur much earlier ($N_r\approx 10^3$).}
\label{wvar_nr}
\end{figure}

@@

In Figure \ref{wvar_nr}, we examine how the degradation in the $w$
constraint depends on the number of simulations used to estimate the
covariance, in the limit of large $N_r$. We find excellent agreement
with the expected scaling (eq.~\ref{ourscaling}) up to $N_r\sim {\rm few}\times10^4$ when using the $\kappa$ power spectrum in the multipole range $\ell\in[100,6000]$. The same behavior is observed when considering the high significance peak counts ($>10\sigma$ for unsmoothed maps and $>5\sigma$ for $1^\prime$ smoothed maps). As the figure shows, around these values of $N_r$ the $\langle\hat{\sigma}^2_w\rangle - \sigma^2_{w,\infty}$ curve
becomes noisy and reaches negative values. This is a clear indication that the $1/N_r$ behavior is broken and a plateau in
$\langle\hat{\sigma}^2_w\rangle$ is reached. The negative values in
the plot are a consequence of the noise in the estimation of this
plateau value (or equivalently in the estimated value of
$\sigma^2_{w,\infty}$). We therefore conclude that a single $N$--body
simulation is sufficient to construct an ensemble of up to a
few$\times 10^4$ mutually independent convergence power spectra. For
$N_r\gg10^4$, the shear realizations can no longer be considered
independent. We should emphasize that the validity of this statement is conditional to the size of the simulation box (which in our case has been chosen to be 240Mpc$/h$) and the range of multipoles $\ell$ used to constrain the parameters. Figure \ref{wvar_nr} shows indeed that, when we try infer $w$ only from large scale modes, $\ell\lesssim250$, the plateau is reached at least an order of magnitude earlier in the number of realizations. In other words, the number of independent power spectra gets lower and lower as we increase the spatial scales of interest. This is due to the fact that, because of the finite box size, the number of independent lens plane shifts (as described in \S~\ref{shearsim}) decreases as the mode size approaches the size of the box.

@@

\item When the feature covariance matrix is measured from simulations,
  parameter constraints are degraded. This degradation is appreciably
  larger than the $O(1/N_r)$ computed by \citep{DodelsonSchneider13}
  when the number of realizations $N_r$ is only a factor of few larger
  than the feature vector size $N_b$.
\item We can recycle a single 240Mpc$/h$ $N$--body simulation to produce an
  ensemble of $O(10^4)$ shear maps whose small scale power spectra and high significance peak counts are statistically independent. The mean feature measured from a shear ensemble, though, could be inaccurate if only one $N$--body simulation is used.
\item As few as one or two independent $N$--body simulations are
  sufficient to forecast $w$ error bars to 1\% accuracy, provided that
  a sufficiently large number $N_r$ of realizations are used to
  measure feature covariances.  In particular, provided that biases
  in the inverse covariance are corrected, percent--level forecasts
  require $N_r\gsim 100 (N_b-N_p)$ realizations.
\item Depending on the feature type used to constrain cosmology, a
  larger number of $N$--body simulations might be needed to measure
  accurate ensemble means to an accuracy corresponding 10\% of the
  statistical error. If this accuracy requirement is relaxed to 50\%
  of the statistical error, we find that as low as $N_s=2$ simulations
  are sufficient for the feature types we consider in this work.
\end{itemize}
%
Future extensions of this work should involve extending our analysis
to a larger set of cosmological parameters, and to more general
feature spaces, such as the ones that characterize non--Gaussian
statistics (e.g. including higher moments of the $\kappa$ field,
Minkowski Functionals, and higher-order $\kappa$ correlators). While
our results are highly encouraging, and suggest that a single $N$--body
simulation can be recycled repeatedly, to produce as many as $10^4$
independent shear power spectra or peak count histograms. In order to
scale our results to large future surveys, such as LSST, it will be
necessary to determine if our findings hold when challenged by larger
and higher-resolution $N$--body simulations \citep{Qcontinuum}.
